---
title: "p8105_hw3_qh2285"
author: "Qianxuan Huang"
date: "2024-10-13"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
library(ggridges)
library(haven)
```

# problem 1

# problem 2

Load and tidy data; exclude participants less than 21 years of age and those with missing demographic data; encode data with reasonable variable classes:
```{r}
participants_df = #data start at row 4
    read_csv("./data/nhanes_covar.csv", na = c("NA", "", "."), skip = 4) |>
  janitor::clean_names() |> 
  mutate(sex = recode(sex, "1" = "male", "2" = "female"),
         education = recode(education, "1" = "Less than high school", 
                            "2" = "High school equivalent", 
                            "3" = "More than high school")) |>
  filter(age >= 21) |>
  drop_na()

accelerometer_df = 
  read_csv("./data/nhanes_accel.csv", na = c("NA", "", ".")) |>
  janitor::clean_names()
```
merge datasets:
```{r}
final_df = participants_df |>
  left_join(accelerometer_df, by = "seqn")
```
reader-friendly table:
number of men and women in each education category:
```{r}
final_df |>
  mutate(education = forcats::fct_relevel(education, c("Less than high school", "High school equivalent", "More than high school"))) |>
  janitor::tabyl(education, sex)
```
create a visualization of the age distributions for men and women in each education category:
```{r}
final_df |>
  mutate(education = forcats::fct_relevel(education, c("Less than high school", "High school equivalent", "More than high school"))) |>#change order
  ggplot(aes(x = age, color = sex)) +
  geom_histogram() +
  labs(title = "Age Distribution by Sex and Education Level",
       x = "Age", y = "Count") +
  theme_minimal() +
  facet_grid(. ~ education) #seperate graphs
```
For female, 28, 23, 59 people in Less than high school, High school equivalent, More than high school; for male they are 27, 35,56.
For people in Less than high school group, distribution and number of female and male are very similar. For people in High school equivalent group, numb of male is bigger than female; and distribution is concentrated at ages very old or young in both sex. For people in More than high school group, number of female and male are very similar, both sex shown left-skewed distribution.





Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant:
```{r}
activity_df = final_df |>
  mutate(education = forcats::fct_relevel(education, c("Less than high school", "High school equivalent", "More than high school"))) |>
  rowwise() |>
  mutate(total_activity = sum(c_across(starts_with("min")))) |>
  select(!(starts_with("min")))
```

Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences:
```{r}
ggplot(activity_df, aes(x = age, y = total_activity, color = sex)) +  
  geom_point() + 
  geom_smooth(se = FALSE) +
  facet_wrap(. ~ education) +  
  labs(title = "Total Activity against Age by Education Level and Gender",
       x = "Age", y = "Total activity")
```
The total activity points don't fit the curve at all because they are too scattered. No trend can be concluded of total activity against age by education level.






Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex:
```{r}
activity_summary = final_df |>
  mutate(education = forcats::fct_relevel(education, c("Less than high school", "High school equivalent", "More than high school"))) |>
  group_by(sex, education) |>
  summarise(across(starts_with("min"), mean, na.rm = TRUE))

# Pivot the dataset to calculate mean of each min
activity_summary_long <- activity_summary |>
  pivot_longer(cols = starts_with("min"), 
               names_to = "minute", 
               names_prefix = "min", 
               values_to = "avg_activity") |>
  mutate(minute = as.numeric(minute))

ggplot(activity_summary_long, aes(x = minute, y = avg_activity, color = sex)) +
  geom_line() +  
  geom_smooth(se = FALSE) +  
  facet_wrap(. ~ education, ncol = 1) + 
  labs(title = "24-hour Average Activity Time Course by Education Level and Sex", x = "Minute", y = "Mean Activity") 
```

Three education level shows same trends throughout the day that low activity at night and increase in the morning. For High school equivalent and More than high school group, female shows higher activity level than male in general, but not in Less than high school group.


# problem 3


```{r}
jan_2020 = 
    read_csv("./data/citibike/Jan 2020 Citi.csv", na = c("NA", "", ".")) |>
    janitor::clean_names() |>
    mutate(month = "January") |>
    mutate(year = 2020)
jan_2024 = 
    read_csv("./data/citibike/Jan 2024 Citi.csv", na = c("NA", "", ".")) |>
    janitor::clean_names() |>
    mutate(month = "January") |>
    mutate(year = 2024)
july_2020 = 
    read_csv("./data/citibike/July 2020 Citi.csv", na = c("NA", "", ".")) |>
    janitor::clean_names() |>
    mutate(month = "July") |>
    mutate(year = 2020)
july_2024 = 
    read_csv("./data/citibike/July 2024 Citi.csv", na = c("NA", "", ".")) |>
    janitor::clean_names() |>
    mutate(month = "July") |>
    mutate(year = 2024)

all_bike = bind_rows(jan_2020, jan_2024, july_2020, july_2024)
```

Produce a reader-friendly table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members:
```{r}
janitor::tabyl(all_bike, year, month, member_casual)
```
For casual riders, in 2020, 984 rides in January, and 5637 in July; in 2024, 2108 rides in January, and 10894 in July.
For city bike member riders, in 2020, 11436 rides in January, and 15411 in July; in 2024, 16753 rides in January, and 36262 in July.



Make a table showing the 5 most popular starting stations for July 2024; include the number of rides originating from these stations:
```{r}
top_starting = july_2024 |>
  group_by(start_station_name) |>
  summarise(total_rides = n()) |>
  arrange(desc(total_rides)) |>
  slice_head(n = 5) 
top_starting
```


Make a plot to investigate the effects of day of the week, month, and year on median ride duration. This plot can include one or more panels, but should facilitate comparison across all variables of interest:
```{r}
median_duration = all_bike |>
    mutate(weekdays = forcats::fct_relevel(weekdays, c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) |>
  group_by(year, month, weekdays) |>
  summarise(median_duration = median(duration, na.rm = TRUE))

(
  median_duration |> 
  filter(year == 2020) |> 
  group_by(month, weekdays) |>
  ggplot(aes(x = weekdays, y = median_duration, group = month, group = weekdays)) +  
  geom_point() +  
  geom_smooth(se = FALSE) +  
  facet_wrap(. ~ month, ncol = 1) +  
  labs(title = "2020", x = "Weekday", y = "Median Ride Duration") +
    ylim(8.5, 16.75)
) + # put 2 plots together
(
  median_duration |> 
  filter(year == 2024) |> 
  group_by(month, weekdays) |>
  ggplot(aes(x = weekdays, y = median_duration, group = month, group = weekdays)) +  
  geom_point() +  
  geom_smooth(se = FALSE) +  
  facet_wrap(. ~ month, ncol = 1) +  
  labs(title = "2024", x = "Weekday", y = "Median Ride Duration") +
    ylim(7.25, 11)
)
```
Comparing between January and July: Median ride duration in July in both year are higher than January, and shows similar trends for weekdays. Especially in July that duration time drop in Friday, the trend keep increase throughout the week otherwise.
Comparing between 2020 and 2024: Median ride duration time are generally much higher in 2020.



For data in 2024, make a figure that shows the impact of month, membership status, and bike type on the distribution of ride duration:
```{r}
all_2024 = bind_rows(jan_2024, july_2024)

all_2024 |>
  group_by(month, member_casual, rideable_type) |>
  ggplot(aes(x = member_casual, y = duration, fill = rideable_type)) +
  geom_boxplot() + 
  facet_wrap(~ month) + 
  labs(x = "Membership Status", y = "Ride Duration", fill = "Bike Type") 
```
According to the box plot, all groups have long tails in long-time ride duration from about 25 to 250. For casual riders, they ride classic bike for a longer time; for city bike members, they ride electric bike for a longer time. Duration of city bike members are more concentrated in shorter time. January and July show similar distribution, wile duration in July is slightly longer.

